`localhierarchy`: an R package to facilitate fitting of global and local Bayesian hierarchical models

### 2025-09-29 paper/paper.qmd
comments evan

Since local models are usually derived from global models, a key challenge in using both for global health estimation is synchronizing codebases and data flows between these approaches. This is particularly important for modeling approaches that involve hierarchical estimation of multiple parameters, where the hierarchical structures may differ across parameters.

@@

This article presents a practical introduction to the \texttt{localhierarchy} package along with illustrative case studies demonstrating its applicability. The article is organized as follows. The next section introduces the modeling approach and workflow used for estimating family planning indicators, motivating the need for the \texttt{localhierarchy} functionality of fitting global and local hierarchical models. The methods section introduces the package' functionality for the estimation of one parameter at the national and subnational level. Finally, we return to the case study of family planning estimation to present illustrative results of the application of the \texttt{localhierarchy} setup in this context.

@@

The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions. Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to as the ratio of modern use to demand. Data on such indicators is available from household surveys and routine data systems. FPET estimates are obtained from fitting a Bayesian model to available data. An illustration of FPET estimates and available data is shown in @fig-localfits for three countries in Western Africa.

@@

Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}. Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the asymptote of the transition of demand from low to high values, and the level of demand in a reference year. @fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model. @fig-asympt shows estimates for the asymptote. These figures show variation in levels among countries in Western Africa, informed by country-specific data. Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition yet; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.

@@

The workflow used in FPET to produce country-level estimates is as follows. Firstly, the full Bayesian model is fitted to a global data base of survey data. Data base compilation is carried out by the United Nations Population Division (ref) and the Track20 project (ref). The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes. Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit.The technical details for this approach are given in the next section. We return to the application of estimating family planning indicators in Section 4.2.

@@

The modeling approach introduced in `localhierarchy` was motivated by application like FPET that depend on hierarchical structures associated with national and subnational level estimation, requiring global and local models to be fit. In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation. We use general notation and a specific example focused on national level estimation. Applications and an extension to subnational estimation are included in the use cases.

@@

Suppose we want to estimate some population parameter of interest $\mu_r$ for a population indexed by $r$. For example, $\mu$ could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. Index $r$ refers to the population in a region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 2: \begin{align*}
  \mu_{r^{(l)}}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma^{(l)}_\theta &\sim N(\mu^{(l-1)}_{r^{(l-1)}}, \sigma^{(l)}), \text{ for } l > 0,
\end{align*} where $\mu_{r^{(l)}}^{(l)}$ ... Here, $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma^{(l)}$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

% --- Level 0 ---
\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};

@@

\begin{alignat}{2}
  \mu_c &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country} &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{r[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{ \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{ \region}^2), \nonumber
\end{alignat} where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$. The parameter $\mu^{(\glob)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality. The parameters $\sigma_{ \country}$, $\sigma_{ \subregion}$, and $\sigma_{ \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.

@@

We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set. A typical workflow for local estimation of national level outcomes is illustrated in Figure 3. For national estimation using the model given by Eq.\~\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model. In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.

```{=latex}
\begin{figure}[h]
\begin{adjustbox}{max size={\textwidth}{\textheight},center}
\begin{tikzpicture}[
  box/.style={rectangle, draw, thick, rounded corners, minimum width=6cm, align=left, inner sep=6pt},
  databox/.style={box, fill=data!20},
  modelbox/.style={box, fill=param!20},
  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
  font=\sffamily
]
% --- Nodes ---
\node[databox] (globaldata) {Global data base of country data};
\node[modelbox, below=1cm of globaldata] (globalmodel) {%
  Global national modeling:\\
  Estimate all parameters in hierarchical model, e.g., \\
\(
  \begin{aligned}
\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \end{aligned}
\)
};
\node[modelbox, below=1cm of globalmodel] (localmodel) {%
  Local national modeling:\\
  Estimate national parameters only, \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
};

@@

$$\mu_c \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{ \country}^2),$$

@@

The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters. Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit. More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ for $k \leq l$ as long as, for a given fixed level $f$, variance parameters at levels $f' < f$ have been fixed. Similarly, the user can fix $\mu$'s for for $k \leq l$ as long as, for a given fixed level $f$, $\mu$s at levels $f' < f$ have been fixed and if the $\sigma$'s for the levels up to and include $f$, i.e., $f' \leq f$, have been fixed.

@@

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

@@

The parameterization used in Stan is a non-centered parameterization. In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
  \mu_r^{(l)} &= \mu^{(l-1)} + \sigma^{(l)} \cdot \gamma_{r^\prime}^{(l-1)},
\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0, s)$, we get \begin{align*}
  \mu_r^{(l)} &= s\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma^{(k)} \cdot \gamma_{r^\prime}^{(k)},
\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parameterization is given by \begin{align*}
  \mu_c =  s_{\glob} \cdot {\gamma}_{\glob} + {\sigma}_{\mu, \region}\cdot {\gamma}_{r[s[c]]]}^{(\region)} + {\sigma}_{\mu, \subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\mu, \country}\cdot\gamma_{c}.
\end{align*}

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c$s: $$(\mu_1, ..., \mu_c) = M\cdot (\eta_1, ..., \eta_{1+R+S+C}),$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\gamma_1 = (s_{\glob}\cdot \gamma_{\glob}, \gamma_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)},..., \gamma_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}$. When estimating a subset of parameters in a local model, fixed values of $\sigma$s and $\gamma$s by point estimates (posterior means) from a global fit. For example, for a national model, we get \begin{align*}
  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{r[s[c]]]}^{(\region)} + 
          \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
          \hat{\sigma}_{ \country} \cdot \gamma_{c}.
\end{align*}

@@

The simulation setup is summarized in Figure 4, panels A and B. Simulation of subnational data is discussed in a later section.

@@

% --- Level 0 ---
\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s^2)$};

% --- Level 1 ---
\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(1)}_1 \sim \mathcal{N}(\mu^{(0)},\sigma_{1}^2)$};
\node[level1, right=of mu11] (mu12) {$\mu^{(1)}_2$};
\node[level1, right=of mu12] (dots1) {$\dots$};
% --- Level 2 ---
\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(2)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{2}^2)$};
\node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
\node[level2, right=of mu22] (dots2) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={yellow!40}{(mu0)}] {};
  \node[background={orange!40}{(mu11)(mu12)(dots1)}] (bg1) {};
  \node[background={green!40}{(mu21)(mu22)(dots2) }] (bg2) {};
\end{pgfonlayer}

% Sigma labels inside background blocks
\node[anchor=north west] at (bg1.east) {$\sigma_{1}$, country};
\node[anchor=north west] at (bg2.east) {$\sigma_{2}$, subnational region};
% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);

@@

% --- Level 1 ---
\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
\node[data, right=of mu11] (mu12) {$y^{(country)}_2$};
\node[data, right=of mu12] (dots1) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
\end{pgfonlayer}

@@

% --- Level 1 ---
\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(1)}_1,\sigma_{y}^2)$};
\node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$};
\node[data, right=of mu12] (dots1) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
\end{pgfonlayer}

@@

% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);
\end{tikzpicture}
\end{adjustbox}}
\caption{Diagram of simulation settings. Top: Hierarchical model for generating country and subnational parameters. Center: Data generating mechanism for national data, illustrated for the first country. Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region.}
\end{figure}
\clearpage
```

Panel A in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$. Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}, \sigma_y^2 \sim N(\mu_{c[i]}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

Simulated data were generated using `simulate_multilevel_data` function. We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$:

```{r}
dat_all <- simulate_multilevel_data(
  n_levels = 2,
  n_units_perlevel = c(30, 20),
  sigma_perlevel = c(0.5, 0.2),
  mu_global = 0,
  add_data = TRUE,
  n_data = 10,
  sigma_y = 0.1,
  add_data1levelup = TRUE
)
dat_nat <- dat_all$data_1levelup
```

In the simulated data, `level1` refers to the country level. The data sets contain the simulated parameters (`eta`) and data (`y`):

@@

Input data are processed using the function `get_geo_unit_index_data`. This function obtains unique hierarchical groupings from the input data and assigns index `c` to each unique grouping:

@@

Excerpts from the Stan model are copied below to introduce the model specification. The same Stan model is used for both global and local models. The first excerpt starts at the end in terms of model specification and concerns the specification of the hierarchical parameters $\mu_c$, using the `localhierarchy` Stan function `get_mu`:

Here we use that $(\mu_1, ..., \mu_c) = M \cdot (\eta_1, ..., \eta_c)$, where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_c)$ refers to the vector with products of standard deviation and $\gamma$ terms. The function `get_mu` calculates the vector of $\mu_c$s from the vector of $\eta$s (`mu_star` in the Stan model) and a decomposition of $M$ into different parts $U$, $V$, $W$. This implementation uses the sparse representation of the model matrix $M$ with Stan functions `csr_matrix_times_vector` and `crs_extract`.

The second Stan model excerpt concerns the specification of the $\eta$s (`mu_star`). This vector is obtained by multiplying the vector of $\gamma$s, which is referred to as `mu_raw` in the Stan model by their respective standard deviation, i.e., the vector containing $s_{\mu, global}$ (labeled `mu_scalarprior_mean` in the Stan model) and the vector of stacked $\sigma$s (labeled `mu_sigma` in the Stan model). To allow for global and local models, additions `_estimate` and `_fixed` are used to allow for splitting the `mu_raw` vector into two parts, one for fixed and one for estimated parameters. Similarly, vector `mu_sigma` is also split into two parts, one part for fixed and one part for estimated parameters. Relevant Stan code is as follows:

In the model code, the data block contains the input data for the $\gamma$s and $\sigma$s. The `localhierarchy` Stan function `get_mu_star` is used to obtain the $\eta$s. Its arguments also include prior parameters `mu_scalarprior_mean` and `mu_scalarprior_sd`, which refer to the (rescaled) prior mean and standard deviation of $\mu^{(\glob)}$. The model block contains the priors for the $\gamma$s and $\sigma_{\mu}$s, given by \begin{align*}
  \mu^{(\glob)} &\sim N(0,1),\\
  \sigma_{\mu} &\sim \text{Half-Normal}(0, s_{\mu}).
\end{align*} All prior parameters are added as input data to the Stan model.

We can use the function `fit_model_localhierarchy` to fit the model as specified so far. The argument `runstep` is used to define the type of model, setting it to `global_national` results in fitting a model to all data without fixing any parameters. Other arguments include the hierarchical level specification, the area, and survey data:

@@

@@

>>>>

A key challenge in the use of local models in global health estimation is synchronizing codebases and data flows between global and local estimation approaches. Synchronizing codebases and data flows is important because the fitting of local models relies on parameter estimates from global models, hence synchronization is important to ensure consistency across global and local models. Synchronization can be particularly challenging for modeling approaches that involve hierarchical estimation of multiple parameters, where the hierarchical structures may differ across parameters.

@@

This article presents a practical introduction to the \texttt{localhierarchy} package along with illustrative case studies demonstrating its applicability. The article is organized as follows. The next section introduces the modeling approach and workflow used for estimating family planning indicators, motivating the need for the \texttt{localhierarchy} functionality of fitting global and local hierarchical models. The methods section introduces the package's functionality for the estimation of one parameter at the national and subnational levels. Finally, we return to the case study of family planning estimation to present illustrative results of the application of the \texttt{localhierarchy} setup in this context.

@@

The Family Planning Estimation Tool (FPET) is widely used to estimate key family planning indicators in countries and regions. Key indicators include modern contraceptive use, demand for family planning, and demand satisfied, referring to the ratio of modern use to demand. Data on such indicators is available from household surveys and routine data systems. FPET estimates are obtained from fitting a Bayesian model to available data. An illustration of FPET estimates and available data is shown in @fig-localfits for three countries in Western Africa.

@@

Given limited country-specific information on some parameters, Bayesian hierarchical models are used to share information between populations, using hierarchical structures such as the one illustrated in Figure \ref{fig-hierstruc}. Specifically for family planning estimation and forecasting, a hierarchical model is used to estimate the asymptote of the transition of demand from low to high values, and the level of demand in a reference year. @fig-level shows estimates of the level in the reference year at the country level for countries in Western Africa, and for the groupings used in the hierarchical model. @fig-asympt shows estimates for the asymptote. These figures show variation in levels among countries in Western Africa, informed by country-specific data. Estimates of asymptotes are similar across countries in Western Africa because countries have not yet reached the end phase of their transition; the asymptotes in Western African countries are informed by the estimate of the asymptote in sub-Saharan Africa through the hierarchical model.

@@

The workflow used in FPET to produce country-level estimates is as follows. Firstly, the full Bayesian model is fitted to a global data base of survey data. Data base compilation is carried out by the United Nations Population Division (ref) and the Track20 project (ref). The global model fit provides information on model parameters, including hierarchical parameters, e.g., estimates of asymptotes. Subsequently, a local model can be used to produce FP estimates based on the data from one country, using information from the global model fit. The technical details for this approach are given in the next section. We return to the application of estimating family planning indicators in Section 4.2.

@@

The modeling approach introduced in `localhierarchy` was motivated by applications like FPET that depend on hierarchical structures associated with national and subnational level estimation, requiring global and local models to be fit. In this section, we introduce the principle setup used for global and local hierarchical models in the context of national estimation. We use general notation and a specific example focused on national level estimation. Applications and an extension to subnational estimation are included in the use cases.

@@

Suppose we want to estimate a population parameter $\mu_r^{(l)}$ of interest, referring to a population health indicator for region $r$ at level $l$ of the hierarchy. For example, the indicator could refer to some population health indicator such as service coverage or morbidity or mortality-related outcomes. For a given level $l$, index $r$ refers to the population in a particular region indexed by $r$, for example, a subnational region, a country, or a group of countries. We assume that $\mu$ is unconstrained (possibly after a transformation has been applied) and consider the following general hierarchical model setup, see Figure 6: 
\begin{align*}
  \mu_{r}^{(l)} \mid \mu^{(l-1)}_{r^\prime}, \sigma_l} &\sim N(\mu^{(l-1)}_{r^\prime}, \sigma_l^2), \text{ for } l > 0,
\end{align*} where $\mu^{(l-1)}_{r^\prime}$ is the parameter value at one level higher in the hierarchy for the region $r^\prime$ that contains unit $r$. The parameter $\sigma_l$ for $l > 0$ describes variability of values of the parameter $\mu$ across geographic units at level $l$ of the hierarchy. At the top level of the hierarchy, $l = 0$, only a single parameter is estimated.

@@

% --- Level 0 ---
\node[level0] (mu0) {$\mu^{(0)} \sim \mathcal{N}(0,s_0^2)$};

@@

\begin{alignat}{2}
  \mu_c^{(\country)}^{(\country)} &\mid \mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2 &&\sim N(\mu_{s[c]}^{(\subregion)}, \sigma_{ \country}^2), \label{eq-countryhier} \\
  \mu_s^{(\subregion)} &\mid \mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2 &&\sim N(\mu_{w[s]}^{(\region)}, \sigma_{ \subregion}^2), \nonumber \\
  \mu_w^{(\region)} &\mid \mu^{(\glob)}, \sigma_{ \region}^2 &&\sim N(\mu^{(\glob)}, \sigma_{ \region}^2), \nonumber
\end{alignat} where $c =1,\dots,C$ refers to the country index, $s[c] = 1, .., S$ to the subregion index of country $c$, and $w[s[c]] = 1, .., W$ to the region index of subregion $s[c]$. The parameter $\mu^{(\glob)}$ is the global mean for the parameter of interest, e.g., a global mean for service coverage or mortality. The parameters $\sigma_{ \country}$, $\sigma_{ \subregion}$, and $\sigma_{ \region}$ describe variability in the parameter of interest across countries, subregions, and regions, respectively.

@@

We use the term local hierarchical models to refer to a hierarchical model in which a subset of parameters have been fixed at posterior mean estimates obtained from a fit of the model to a larger data set. A typical workflow for local estimation of national level outcomes is illustrated in Figure 7. For national estimation using the model given by Eq.\ref{eq-countryhier}, the user starts by fitting a global model, using a global data base and estimating all parameters in a hierarchical model. In a local model, data from one country are used and parameters that are not country-specific are fixed using parameter estimates from the global model fit.

```{=latex}
\begin{figure}[h]
\begin{adjustbox}{max size={\textwidth}{\textheight},center}
\begin{tikzpicture}[
  box/.style={rectangle, draw, thick, rounded corners, minimum width=6cm, align=left, inner sep=6pt},
  databox/.style={box, fill=data!20},
  modelbox/.style={box, fill=param!20},
  arrowdata/.style={-{Latex[length=3mm]}, thick, data},
  arrowparam/.style={-{Latex[length=3mm]}, thick, param!80},
  font=\sffamily
]
% --- Nodes ---
\node[databox] (globaldata) {Global data base of country data};
\node[modelbox, below=1cm of globaldata] (globalmodel) {%
  Global national modeling:\\
  Estimate all parameters in hierarchical model, e.g., \\
\(
  \begin{aligned}
\mu_w^{(region)} &\sim N(\mu^{(global)}, \sigma_{ region}^2),\\
\mu_s^{(subregion)} &\sim N(\mu_{w[s]}^{(region)}, \sigma_{ subregion}^2), \\
\mu_c^{(country)} &\sim N(\mu_{s[c]}^{(subregion)}, \sigma_{ country}^2),  \end{aligned}
\)
};
\node[modelbox, below=1cm of globalmodel] (localmodel) {%
  Local national modeling:\\
  Estimate national parameters only, \\
  fix parameter estimates at higher levels of the hierarchy \\i.e., $\mu_c^{(country)}$ with $\mu_c^{(country)} \sim N(\hat{\mu}_{s[c]}^{(subregion)}, \hat{\sigma}_{ country}^2).$
};

@@

$$\mu_c^{(\country)} \sim N(\hat{\mu}_{s[c]}^{(\subregion)}, \hat{\sigma}_{ \country}^2),$$

@@

The `localhierachy` package provides a general implementation of local models by facilitating the fixing of subsets of parameters. Continuing with the general notation introduced above, in the package, a local model for one parameter of level $l$ is derived by fixing $\sigma^{(k)}$ for $k=1, ..., l$ and fixing $\mu_{r\prime}{(k)}$ for $k = 1, .., l-1$ at posterior means derived from a global fit. More generally, if a model was fitted using data up to level $l$, the user can create a local model by fixing $\sigma^{(k)}$ at all levels $k$ up to some chosen level $K_{\sigma}$. Similarly, the $\mu^{(k)}$ parameters can be fixed at all levels $k$ up to a level $K_{\mu} \leq K_{\sigma}$.

@@

The package includes a user-friendly specification of hierarchical model structures, and associated specification of fixed parameters for local runs. A worked example is given in the first use case. Briefly, if a user wants to specify a hierarchical model like the one introduced above, with the levels world, regions, subregion, and country, <!-- country parameters $\mu_c^{(\country)} =  s_{global}\cdot \gamma_{global} + \sigma_{ region}\cdot \gamma_{r[s[c]]]}^{(
egion)} + \sigma_{ subregion}\cdot\gamma_{s[c]}^{(\subregion)} + \sigma_{ country}\cdot\gamma_{ c}$,  --> using a dataset with information on these groupings, with naming `region`, `subregion`, and `country`, in the R functions, the user can specify the hierarchical model as follows \[Q for herb: refs/credit to be given for this set up to other work?\]:

@@

The parameterization used in Stan is a non-centered parameterization. In general notation, for $l>0$, non-centered parameterization refers to writing mean parameters as deviations from higher-level group means: \begin{align*}
  \mu_r^{(l)} &= \mu^{(l-1)} + \sigma_l \cdot \gamma_{r}^{(l)},
\end{align*} where $\gamma_r^{(l)} \sim N(0, 1)$. Applied sequentially, and if $\mu^{(0)} \sim N(0,s_{\glob})$ with prior standard deviation $s_{\glob}$, we get \begin{align*}
  \mu_r^{(l)} &= s_{\glob}\cdot\gamma^{(0)} + \sum_{k=1}^{l} \sigma_k \cdot \gamma_{r^\prime}^{(k)},
\end{align*} where $r^\prime$ now refers more generally to the index of the unit containing the lower-level group. For the specific example of country parameters, the non-centered parameterization is given by \begin{align*}
  \mu_c^{(\country)} =  s_{\glob} \cdot {\gamma}^{(\glob)} + {\sigma}_{\region}\cdot {\gamma}_{w[s[c]]]}^{(\region)} + {\sigma}_{\subregion}\cdot{\gamma}_{s[c]}^{(\subregion)} + {\sigma}_{\country}\cdot\gamma_{c}^{(\country)}.
\end{align*}

Implementation is based on a model matrix. For our model given by the equation above, the following matrix multiplication is used to obtain the vector of $\mu_c^{(\country)}$s: $$(\mu_1^{(\country)}, ..., \mu_c^{(\country)}) = M\cdot (\eta_1, ..., \eta_{1+R+S+C})',$$ where matrix $M$ is a sparse matrix with 1s and 0s, and $(\eta_1, ..., \eta_{1+R+S+C})$ refers to the vector of the product of the standard deviation terms and $\gamma$s, i.e., $\eta_1 = s_{\glob}\cdot \gamma_{\glob}, \eta_2 = \sigma_{ \region} \cdot \gamma_{ 1}^{(\region)}, ..., \eta_{1+W+S+C} = \sigma_{ \country}\cdot\gamma_{ C}^{(\country)}$. When estimating a subset of parameters in a local model, the values of $\sigma$s and $\gamma$s are fixed to point estimates (posterior means) from a global fit. For example, for a national model, we get \begin{align*}
  \mu_c = s_{\glob} \cdot \hat{\gamma}_{\glob} + 
          \hat{\sigma}_{ \region} \cdot \hat{\gamma}_{w[s[c]]]}^{(\region)} + 
          \hat{\sigma}_{ \subregion} \cdot \hat{\gamma}_{s[c]}^{(\subregion)} + 
          \hat{\sigma}_{ \country} \cdot \gamma_{c}^{(\country)}.
\end{align*}

@@

The simulation setup is summarized in Figure 8, panels A and B. Simulation of subnational data is discussed in a later section.

@@

% --- Level 0 ---
\node[level0] (mu0) {$\mu^{(\glob)} \sim \mathcal{N}(0,s_\glob^2)$};

% --- Level 1 ---
\node[level1, below left=1.5cm and 3cm of mu0] (mu11) {$\mu^{(\country)}_1 \sim \mathcal{N}(\mu^{(\glob)},\sigma_{\country}^2)$};
\node[level1, right=of mu11] (mu12) {$\mu^{(\country)}_2$};
\node[level1, right=of mu12] (dots1) {$\dots$};
% --- Level 2 ---
\node[level2, below left=1.5cm and 0cm of mu11] (mu21) {$\mu^{(\subnat)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{\subnat}^2)$};
\node[level2, right=of mu21] (mu22) {$\mu^{(2)}_2$};
\node[level2, right=of mu22] (dots2) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={yellow!40}{(mu0)}] {};
  \node[background={orange!40}{(mu11)(mu12)(dots1)}] (bg1) {};
  \node[background={green!40}{(mu21)(mu22)(dots2) }] (bg2) {};
\end{pgfonlayer}

% Sigma labels inside background blocks
\node[anchor=north west] at (bg1.east) {$\sigma_{\country}$, country};
\node[anchor=north west] at (bg2.east) {$\sigma_{\subnat}$, subnational region};
% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);

@@

% --- Level 1 ---
\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(country)}_1 \sim \mathcal{N}(\mu^{(\country)}_1,\sigma_{y}^2)$};
\node[data, right=of mu11] (mu12) {$y^{(country)}_2$};
\node[data, right=of mu12] (dots1) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
\end{pgfonlayer}

@@

% --- Level 1 ---
\node[data, below left=1.5cm and 3cm of mu0] (mu11) {$y^{(subnat)}_1 \sim \mathcal{N}(\mu^{(\subnat)}_1,\sigma_{y}^2)$};
\node[data, right=of mu11] (mu12) {$y^{(subnat)}_2$};
\node[data, right=of mu12] (dots1) {$\dots$};
% --- Background blocks ---
\begin{pgfonlayer}{background}
  \node[background={gray!40}{(mu11)(mu12)(dots1)}] (bg1) {};
\end{pgfonlayer}

@@

% --- Arrows ---
\draw[arrow] (mu0.south) -- (mu11.north);
\draw[arrow] (mu0.south) -- (mu12.north);
\draw[arrow] (mu0.south) -- (dots1.north);
\end{tikzpicture}
\end{adjustbox}}
\caption{Diagram of simulation settings. Top: Hierarchical model for generating country and subnational parameters. Center: Data generating mechanism for national data, illustrated for the first country. Bottom: Data generating mechanism for subnational data, illustrated for the first subnational region (used only in Section 5).}
\end{figure}
\clearpage
```

Panel A in Figure 4 indicates the hierarchical model considered. For national level estimation, we consider countries to be grouped in the world, i.e., $\mu_c^{(\country)}|\mu^{(\glob)}, \sigma_{ \country} \sim N(\mu^{(\glob)}, \sigma_{ \country}^2)$. Panel B introduces the likelihood function for national data, given by a normal likelihood with constant variance: $$y_i^{(\country)}|\mu_{c[i]}^{(\country)}, \sigma_y^2 \sim N(\mu_{c[i]}^{(\country)}, \sigma_y^2),$$ where $c[i]$ refers to the country of country observation $i$. We chose simple normal densities as data generating mechanisms in the simulation, to focus the modeling on the hierarchical model and the `localhierarchy` functionality as opposed to the data model aspects of modeling. In practice, the data model can be set by the user to reflect actual data generating mechanisms, for example, by using a binomial likelihood, considering transformed outcomes, or accounting for sampling errors for observations that were obtained from a complex survey design.

Simulated data were generated using `simulate_multilevel_data` function. We simulate a set of model parameters, setting $\mu^{(\glob)} = 0$, and $\sigma_{\country} = 0.5$, creating 30 countries, and national data, using $\sigma_y = 0.1$. The function call is as follows (note that the function also generates parameters and data at the subnational level, these are used in Section 5):

```{r}
dat_all <- simulate_multilevel_data(
  n_levels = 2,
  n_units_perlevel = c(30, 20),
  sigma_perlevel = c(0.5, 0.2), # national and subnational level
  mu_global = 0,
  add_data = TRUE,
  n_data = 10,
  sigma_y = 0.1, # used for national and subnational data
  add_data1levelup = TRUE
)
dat_nat <- dat_all$data_1levelup
```

In the simulated data, `level1` refers to the country level. The data sets contain the simulated parameters (`eta`; the deviation away from the global mean) and data (`y`):

@@

Input data are processed using the function `get_geo_unit_index_data`. This function obtains unique hierarchical groupings from the input data and assigns an index `c` to each unique grouping:

@@

We can use the function `fit_model_localhierarchy` to fit the model as specified so far. The argument `runstep` is used to define the type of model, setting it to `"global_national"` results in fitting a model to all data without fixing any parameters. Other arguments include the hierarchical level specification, the area, and survey data:

@@

When doing model fitting, the same Stan model is used for both global and local models. Excerpts from the Stan model are included in Appendix 1.

@@

# Appendix
